---
title: "IRCA Exposure and Composition of Admissions"
subtitle: "How denominator choice changes a headline admissions-composition metric"
author: "Kai A. Sanchez P."
date: " February 2026"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{tabularray}
  - \UseTblrLibrary{booktabs}
  - \usepackage{siunitx}
  - \sisetup{detect-all}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{graphicx}
  - \usepackage{adjustbox}
  - \usepackage{capt-of}
  - \usepackage[none]{hyphenat}
  - \usepackage{pdflscape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center",
  out.width = "0.85\\linewidth",
  fig.pos = "H"
)

fig_dir <- "../outputs/figures"
tab_dir <- "../outputs/tables"

fig <- function(x) file.path(fig_dir, x)
tab <- function(x) file.path(tab_dir, x)

tab_tex <- function(name, caption = NULL){
  x <- readLines(tab(name), warn = FALSE)

  drop <- grepl("^\\\\begin\\{table\\*?\\}(\\[.*\\])?\\s*$", x) |
          grepl("^\\\\end\\{table\\*?\\}\\s*$", x) |
          grepl("^\\\\centering\\s*$", x) |
          grepl("^\\\\caption\\{", x) |
          grepl("^\\\\label\\{", x)

  x <- x[!drop]
  body <- paste(x, collapse = "\n")

  cat("\n\\par\\noindent\n")
  if (!is.null(caption)) cat("\\captionof{table}{", caption, "}\n", sep = "")

  cat("\\begingroup\\footnotesize\n")
  cat("\\begin{adjustbox}{max width=\\linewidth}\n")
  cat(body, "\n")
  cat("\\end{adjustbox}\n")
  cat("\\endgroup\n")
  cat("\\par\n\n")
}

safe_input_table <- function(name, caption = NULL){
  f <- tab(name)
  if (!file.exists(f)) {
    cat("\n\\textbf{[MISSING TABLE: ", name, "]}\\par\n", sep = "")
    return(invisible(NULL))
  }
  tab_tex(name, caption = caption)
}

include_fig_safe <- function(name){
  f <- fig(name)
  if (!file.exists(f)) {
    cat(paste0("**[MISSING FIGURE: ", name, "]**"))
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}

stopifnot(dir.exists(fig_dir), dir.exists(tab_dir))
```

# Executive Summary

This project builds a reproducible country–year panel of 19 Latin American origin countries from 1982 to 2000 to study whether higher IRCA legalization exposure is associated with changes in the recorded occupational composition of admissions during 1989–1992.

The main finding is straightforward:

1. A decline in the recorded blue-collar share appears only when the denominator excludes “no occupation.

2. The headline blue-collar share decline is largely a denominator and reporting result, not clear evidence of a collapse in blue-collar admissions.

This is therefore best understood as a data quality and KPI interpretation case study, not a causal evaluation of IRCA. The report shows how denominator choice, missingness, and administrative reporting shifts can materially alter a commonly used composition metric.

# Why This Matters for a Data Analyst

This project is designed as a practical analytics exercise, not just a policy-history analysis.

A reporting or operations analytics team could easily monitor a metric like “blue-collar share” and conclude that composition changed sharply during a legalization surge. But that interpretation can be wrong if the metric depends on an unstable denominator.

The broader analyst lesson is:

> When administrative data quality or coding coverage shifts, percentage-based KPIs can move for mechanical reasons rather than substantive ones.

That makes this a useful portfolio example in:

- administrative data analysis

- dashboard KPI design

- denominator sensitivity auditing

- robustness analysis under imperfect measurement

# Research Question

The analytical question is:

**Do origin countries with higher IRCA legalization exposure show a true decline in blue-collar admissions share during 1989–1992, or does the result mainly reflect denominator and reporting changes?**

This framing is more useful than simply asking whether IRCA “reduced” blue-collar shares, because the key empirical issue in this dataset is how the occupational denominator is constructed.

# Data and Exposure Design

The analysis uses author-compiled spreadsheets based on publicly available INS Statistical Yearbooks. The report uses four harmonized country–year inputs:

- occupation

- class of admission

- gender/age

- adjustments versus new arrivals

The sample covers 19 Latin American origin countries from 1982 to 2000.

## Exposure measure

Country-level IRCA exposure is defined as:

\[
\text{Exposure}=\frac{\text{IRCA legalizations (1989–1992)}}{\text{Pre-IRCA admissions (1982–1985)}}.
\]

Countries with larger legalization surges relative to their earlier admissions base are treated as more highly exposed.

## Design

The core models are descriptive exposure-gradient panel regressions with:

- country fixed effects

- year fixed effects

- clustered standard errors by origin country

The estimates are not interpreted causally. Timing diagnostics later in the report show pre-period differences, so the regressions are used as descriptive summaries rather than identification-based treatment effects.

# Main Result: The Headline Depends on the Denominator

The central portfolio result is the same-sample comparison below. It asks whether the blue-collar share result survives when the denominator is expanded to include records with no occupation.



```{r, results='asis'}
safe_input_table(
  "tab_MAIN_net_vs_full_denom.tex",
  caption = "Main result: blue-collar share using net versus full denominator on the same sample."
)
```

This is the most important table in the report.

Using the net denominator (known occupations only), higher IRCA exposure is associated with a decline in the blue-collar share during 1989–1992. But when the full denominator includes “no occupation,” the association becomes much smaller.

That contrast strongly suggests that the headline result is sensitive to occupation coverage and denominator construction. In other words, the apparent blue-collar share decline is not enough by itself to conclude that blue-collar admissions truly collapsed.

# Mechanism: Numerator vs Denominator Growth

A falling share can reflect either:

- a declining numerator, or

- a denominator that grows faster than the numerator

To distinguish between those explanations, the project decomposes the result into blue-collar counts and denominator counts.




```{r fig-decomp-levels, fig.cap="Decomposition in levels: IRCA × Post associations on blue-collar counts and denominators.", out.width="0.92\\linewidth", fig.pos="H"}
include_fig_safe("fig_decomp_levels_only.png")
```

The decomposition supports a denominator-based interpretation. The blue-collar numerator does not show evidence of a simple collapse. Instead, the denominator grows strongly during high-exposure surge years, which is consistent with the idea that the observed share decline is largely mechanical.

That is the key analytical move in the project: not stopping at the share, but asking what changed underneath it.

# Denominator Channel Evidence: “No Occupation” Share

A natural way to test whether denominator mechanics matter is to examine the behavior of the “no occupation” category itself.



```{r fig-noocc-trends, fig.cap="No-occupation share over time by IRCA exposure group.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_noocc_trends_by_exposure.png")
```



The figure shows that more highly exposed origin countries experienced a larger rise in the no-occupation share around the IRCA surge years. That reinforces the interpretation that occupation coverage changed at the same time the net-denominator blue-collar share was falling.

For an analytics audience, this is the main operational takeaway: the blue-collar KPI should not be interpreted in isolation. It should be monitored jointly with the full-denominator share and the no-occupation share.

# Robustness: Is the Result Driven by One Country?

A common concern in small-country panels is that one country may dominate the estimated relationship. The leave-one-country-out exercise checks that.


```{r fig-loo, fig.cap="Leave-one-country-out influence on the baseline net-denominator coefficient.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_loo_blue_net.png")
```


The result is not driven by any single origin country. The coefficient moves as countries are removed one at a time, but the overall pattern remains intact.

This matters because it shows that the denominator-sensitivity story is not just a Mexico artifact or a single-country anomaly.

# Small-Cluster Inference

Because the panel has only 19 origin-country clusters, conventional clustered standard errors may be fragile. To address that, the project adds a wild cluster bootstrap check for the baseline net-denominator specification.


```{r, results='asis'}
safe_input_table(
  "tab_wildboot_blue_net.tex",
  caption = "Wild cluster bootstrap inference for the baseline net-denominator specification."
)
```

The bootstrap is estimated using an equivalent country-and-year dummy regression because of package compatibility constraints. It serves as an inference check rather than a separate substantive model.

# Magnitudes

To make the results more interpretable, the project reports standardized magnitudes under both denominator definitions.




```{r, results='asis'}
safe_input_table(
  "tab_magnitudes_blue_NET.tex",
  caption = "Standardized effect sizes for blue-collar share using the net denominator."
)
safe_input_table(
  "tab_magnitudes_blue_FULLDEN.tex",
  caption = "Standardized effect sizes for blue-collar share using the full denominator."
)
```



The contrast in magnitudes reinforces the same conclusion as the main result table: the apparent blue-collar decline is materially smaller once denominator coverage is handled more conservatively.

# Reporting Recommendation for Analysts

For dashboarding or operations monitoring, the report recommends presenting three metrics together:

1. Blue share (net denominator) 

2. Blue share (full denominator) 

3. No-occupation share

That combination makes it easier to tell whether a change in the headline composition metric reflects:

- genuine compositional change, or

- reporting/coding shifts in the underlying records

In practical terms, this is the best stakeholder-facing takeaway of the project.

# Interpretation and Limits

This is a descriptive exposure-gradient analysis, not a causal design.

That distinction matters because the event-study timing checks show pre-period differences. More highly exposed origin countries were already on somewhat different trajectories before the IRCA surge, so the interaction coefficients should be interpreted as summary associations rather than causal treatment effects.

That limitation does not weaken the portfolio value of the project. In fact, it strengthens it: the report shows that the analyst did not confuse statistical output with identification.

# Main Takeaways

This project produces four main takeaways:

1. The headline blue-collar share result is highly sensitive to denominator choice.

2. The underlying mechanism is better described as denominator/reporting expansion than as a pure decline in blue-collar counts.

3. The no-occupation category is central to interpreting occupational composition during surge years.

4. Administrative-data KPIs should be audited for denominator instability before being used for substantive conclusions.

# Portfolio Takeaway

As a portfolio project, this case study demonstrates the ability to:

- build a reproducible panel from raw administrative-style spreadsheets

- define and test a country-level exposure measure

- identify denominator sensitivity in a commonly used KPI

- decompose a share into numerator and denominator channels

- run and interpret robustness checks without overstating causality

- translate a technical result into a stakeholder-facing measurement lesson

That is the main reason this project works well as data analyst portfolio material: it shows judgment, not just model execution.

# How to Run

1. Put the raw Excel inputs in `data_raw/`

2. Open the project file

3. Run `Rscript run.R` or source `run.R`

4. Render `paper.Rmd`

Outputs are written to:

- `outputs/tables/`

- `outputs/figures/`

- `outputs/logs/`

- `data_clean/panel.rds`

# Sources

- U.S. Immigration and Naturalization Service (INS). *Statistical Yearbook of the Immigration and Naturalization Service*, various years.

# Appendix

## Appendix A: Portfolio dashboard

```{r, fig-dashboard, fig.cap="Dashboard of shares, no-occupation share, and counts by IRCA exposure group.", out.width="0.95\\linewidth", fig.pos="H"}
include_fig_safe("fig_PORTFOLIO_dashboard.png")
```


## Appendix B: Data quality and coverage

```{r, results='asis'}
safe_input_table(
  "tab_missing_overall.tex",
  caption = "Missingness by variable at the country–year level."
)
safe_input_table(
  "tab_coverage_country.tex",
  caption = "Coverage by origin country for key variables."
)
```

## Appendix C: Baseline regression panels

```{r, results='asis'}
safe_input_table(
  "tab_main_total_A.tex",
  caption = "Baseline TWFE (Total): occupation shares."
)
safe_input_table(
  "tab_main_total_B.tex",
  caption = "Baseline TWFE (Total): no-occupation share and class-of-admission shares."
)
safe_input_table(
  "tab_main_net_A.tex",
  caption = "Baseline TWFE (Net): occupation shares controlling for adjustment share."
)
safe_input_table(
  "tab_main_net_B.tex",
  caption = "Baseline TWFE (Net): no-occupation share and class-of-admission shares controlling for adjustment share."
)
```


## Appendix D: Additional robustness

```{r, results='asis'}
safe_input_table(
  "tab_spec_compare_blue.tex",
  caption = "Specification comparison for blue-collar share (net denominator)."
)
safe_input_table(
  "tab_blue_trends_robustness.tex",
  caption = "Robustness to origin-specific linear trends."
)
safe_input_table(
  "tab_robust_extra_blue.tex",
  caption = "Additional sensitivity checks for blue-collar share."
)
safe_input_table(
  "tab_weights_blue.tex",
  caption = "Weighted specifications."
)
safe_input_table(
  "tab_balanced_blue.tex",
  caption = "Balanced-panel restriction."
)
safe_input_table(
  "tab_alt_intensity_blue.tex",
  caption = "Alternative exposure construction and Mexico exclusion."
)
```

## Appendix E: Decomposition tables and log specification

```{r  results='asis'}
safe_input_table(
  "tab_decomp_levels_A.tex",
  caption = "Decomposition in levels."
)
safe_input_table(
  "tab_decomp_levels_B.tex",
  caption = "Decomposition in log(1+x)."
)
```

```{r fig-decomp-logs, fig.cap="Decomposition in log(1+x): proportional associations on numerator and denominators.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_decomp_logs_only.png")

```


## Appendix F: Timing diagnostics

```{r fig-event-study, fig.cap="Event-study: exposure-by-year associations for blue-collar share (relative to 1988).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_es_blue_total.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_wald_pretrends_es_blue_total.tex",
  caption = "Wald test of joint significance for pre-1988 event-study coefficients."
)

```

The pretrend diagnostic is important for interpretation. It indicates that higher-exposure countries were already evolving differently before the IRCA surge, which is why the main estimates are presented as descriptive rather than causal.

```{r fig-pretrend-groups, fig.cap="Pre-IRCA group means for blue-collar share by exposure group, 1982–1988.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_pretrend_groups.png")
```

```{r fig-early-late, fig.cap="Early versus late post-period exposure-gradient associations for blue-collar share.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_early_late_blue.png")
```


## Appendix G: Additional exploratory outputs



```{r fig-mechanism-scan, fig.cap="Exposure × Post associations across outcomes.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_mechanisms_coefplot.png")
```


## Appendix H: QA artifact

```{r fig-residuals-year, fig.cap="Mean residual by year in the baseline net-denominator specification.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_residuals_by_year_blue.png")
```

