---
title: "IRCA Exposure and Composition of Admissions"
subtitle: "A descriptive exposure-gradient analysis, 1982–2000"
author: "Kai A. Sanchez P."
date: " February 2026"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{tabularray}
  - \UseTblrLibrary{booktabs}
  - \usepackage{siunitx}
  - \sisetup{detect-all}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{graphicx}
  - \usepackage{adjustbox}
  - \usepackage{capt-of}
  - \usepackage[none]{hyphenat}
  - \usepackage{pdflscape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center",
  out.width = "0.85\\linewidth",
  fig.pos = "H"
)

fig_dir <- "../outputs/figures"
tab_dir <- "../outputs/tables"

fig <- function(x) file.path(fig_dir, x)
tab <- function(x) file.path(tab_dir, x)

tab_tex <- function(name, caption = NULL){
  x <- readLines(tab(name), warn = FALSE)

  drop <- grepl("^\\\\begin\\{table\\*?\\}(\\[.*\\])?\\s*$", x) |
          grepl("^\\\\end\\{table\\*?\\}\\s*$", x) |
          grepl("^\\\\centering\\s*$", x) |
          grepl("^\\\\caption\\{", x) |
          grepl("^\\\\label\\{", x)

  x <- x[!drop]
  body <- paste(x, collapse = "\n")

  is_tabularray <- any(grepl("\\\\begin\\{talltblr\\}", x)) || any(grepl("\\\\begin\\{tblr\\}", x))

  cat("\n\\par\\noindent\n")
  if (!is.null(caption)) cat("\\captionof{table}{", caption, "}\n", sep = "")

  cat("\\begingroup\\footnotesize\n")
  cat("\\begin{adjustbox}{max width=\\linewidth}\n")
  cat(body, "\n")
  cat("\\end{adjustbox}\n")
  cat("\\endgroup\n")

  cat("\\par\n\n")
}

safe_input_table <- function(name, caption = NULL){
  f <- tab(name)
  if (!file.exists(f)) {
    cat("\n\\textbf{[MISSING TABLE: ", name, "]}\\par\n", sep = "")
    return(invisible(NULL))
  }
  tab_tex(name, caption = caption)
}

include_fig_safe <- function(name){
  f <- fig(name)
  if (!file.exists(f)) {
    cat(paste0("**[MISSING FIGURE: ", name, "]**"))
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}

stopifnot(dir.exists(fig_dir), dir.exists(tab_dir))
```

# Executive Summary

This project builds a reproducible country–year panel (19 Latin American origins, 1982–2000) to document whether IRCA legalization intensity (1989–1992) is associated with changes in the recorded occupational composition of admissions.

**What I find**: The blue-collar “share decline” during the IRCA surge depends on how you define the denominator. Using known occupations only, higher exposure is linked to a sizeable drop in the blue-collar share. When I include “no occupation” in the denominator, the association largely disappears—suggesting the KPI is driven by reporting/denominator shifts during surge years.

**How to read the estimates**: This is a descriptive exposure-gradient analysis, not a causal design. Event-study timing checks show pre-period differences, so I use the regression results as summary associations and focus on robustness plus numerator/denominator decomposition.

**Reproducibility**: All tables/figures are generated from the raw Excel inputs via run.R (panel build → QA → models → exports).

# Portfolio Exhibits

If you skim: start with Figure 1 → Table 1 → Figure 2, then use Figures 3–4 as diagnostics.

```{r fig-dashboard, fig.cap="Shares, no-occupation share, and counts by IRCA exposure group.", out.width="0.95\\linewidth", fig.pos="H"}
include_fig_safe("fig_PORTFOLIO_dashboard.png")
```



# Research Question and Data

The input spreadsheets in data_raw/ were compiled by the author from publicly available INS Statistical Yearbooks. The report uses only the series required for this analysis (occupation, class of admission, gender/age, adjustments vs new arrivals). See Sources section for publication references.

**Question**: Do origin countries with higher IRCA legalization intensity display larger changes in the occupational composition of admissions during 1989–1992?

Exposure (country-level):

\[
\text{Exposure}=\frac{\text{IRCA legalizations (1989–1992)}}{\text{Pre-IRCA admissions (1982–1985)}}.
\]

**Data sources**: INS Statistical Yearbooks / OHSS tables on occupation, class of admission, gender/age, and adjustments vs new arrivals.

**Design**: descriptive exposure-gradient panel regressions with country and year fixed effects; clustered SE at the origin-country level.




# Main Result: Denominator Sensitivity

The main portfolio result is that the blue-share association depends strongly on whether “no occupation” is excluded (net denominator) or included (full denominator). To avoid sample-driven artifacts, the headline comparison uses the same set of country–years for both outcomes.

```{r, results='asis'}
safe_input_table(
  "tab_MAIN_net_vs_full_denom.tex",
  caption = "Blue share with Net vs Full denominator (same sample)."
)
```

Given the timing diagnostics (Appendix E), interpret this coefficient as a descriptive summary of the exposure gradient—not a causal estimate. In practice: the headline shift is largely explained by changes in occupation coverage (how many records have usable occupation codes), so the dashboard should always pair the net share with the full-denominator share and the no-occupation share.

**Practical implication**: during large legalization/processing surges, composition metrics based on “known occupations only” can move sharply even when the underlying category counts rise — because denominators (including missing/unknown occupation mass) change. For example, if more records are processed without occupation coding during surge years, the ‘known-occupation’ denominator changes even if blue-collar counts rise.




# Mechanism: Numerator vs Denominator Growth

To understand why a share can fall even when counts rise, I decompose the association into numerator and denominator components (levels). The pattern is consistent with denominator growth outpacing numerator growth, rather than a collapse in blue-collar admissions.




```{r fig-decomp-levels, fig.cap="IRCA × Post associations on numerator (blue counts) and denominators (levels).", out.width="0.92\\linewidth", fig.pos="H"}
include_fig_safe("fig_decomp_levels_only.png")
```



# Denominator Channel Evidence: “No Occupation”

A key diagnostic for denominator mechanics is whether the “no occupation” category shifts differentially by exposure during the surge.



```{r fig-noocc-trends, fig.cap="No-occupation share over time by IRCA exposure group.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_noocc_trends_by_exposure.png")
```


# Robustness

The negative net-denominator association is not driven by a single origin country.


```{r fig-loo, fig.cap="Leave-one-country-out influence on the IRCA × Post coefficient (net denominator).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_loo_blue_net.png")
```


**Small-cluster inference note**: with 19 clusters, conventional clustered SE can be fragile. I therefore include a wild cluster bootstrap p-value for the baseline specification (Appendix contains full robustness tables).

```{r, results='asis'}
safe_input_table(
  "tab_wildboot_blue_net.tex",
  caption = "Small-cluster inference add-on: Wild cluster bootstrap test for IRCA × Post (baseline net specification)."
)
```

Wild bootstrap computed on an equivalent country+year dummy specification (same model, different implementation) due to package compatibility.

# Magnitudes

Because the headline depends on denominator definition, magnitudes are reported for both constructions.




```{r, results='asis'}
safe_input_table("tab_magnitudes_blue_NET.tex",
  caption = "Standardized effect sizes for net-denominator blue share (same-sample comparison)."
)
safe_input_table("tab_magnitudes_blue_FULLDEN.tex",
  caption = "Standardized effect sizes for full-denominator blue share (same-sample comparison)."
)
```



# Stakeholder and Decision context (DA framing)

## Intended stakeholder and decision use

**Who this is for**: an admissions reporting / operations analytics team (e.g., OHSS-style reporting or an internal USCIS analytics group).

**Decision use:** during surge windows (large legalization/processing events), this analysis supports:

- **Monitoring:** detect whether composition changes reflect real shifts vs. reporting/denominator changes.

- **Quality assurance:** flag periods when “no occupation” rises and occupation coding may be less consistent.

- **Capacity planning:** anticipate downstream staffing needs if denominator growth reflects processing/recording expansion.

## Metrics specifications (dashboard KPIs)

All shares are in $[0,1]$ and should be displayed as percentage points in dashboards. For each origin country \(c\) and year \(t\):

- **Blue share (net denominator)**  
  \[
  \text{BlueShare}_{ct}^{net}=\frac{\text{BlueCollarNonFarm}_{ct}}{\text{OccKnown}_{ct}}
  \]
  where  
  \[
  \text{OccKnown}_{ct}=\text{WhiteHigh}_{ct}+\text{WhiteSupport}_{ct}+\text{BlueCollarNonFarm}_{ct}+\text{Agriculture}_{ct}+\text{Services}_{ct}.
  \]

- **Blue share (full denominator)**  
  \[
  \text{BlueShare}_{ct}^{full}=\frac{\text{BlueCollarNonFarm}_{ct}}{\text{OccKnown}_{ct}+\text{NoOccupation}_{ct}}.
  \]

- **No-occupation share (full denominator)**  
  \[
  \text{NoOccShare}_{ct}=\frac{\text{NoOccupation}_{ct}}{\text{OccKnown}_{ct}+\text{NoOccupation}_{ct}}.
  \]

In the pipeline (run.R), these correspond to blue_share (net), blue_fullden (full), and noocc_share.

**Why report all three together:** 

\(\text{BlueShare}^{net}\) can move sharply if the **denominator changes** (e.g., classification shifts into “no occupation” or changes in reporting coverage). Reporting \(\text{BlueShare}^{full}\) and \(\text{NoOccShare}\) alongside the net share helps distinguish **true compositional change** from **measurement/reporting mechanics**.

# How to Run

1.  Put the raw Excel inputs in data_raw/.

2.  Open `irca-composition.Rproj`  

3.  Run `Rscript run.R` (or Source `run.R`) 

4.  Render `paper.Rmd`

5.  Outputs are written to:

  - outputs/tables/, outputs/figures/, outputs/logs/, and data_clean/panel.rds


# Sources

- U.S. Immigration and Naturalization Service (INS). *Statistical Yearbook of the Immigration and Naturalization Service*, various years (occupation and admissions tables).



# Appendix

## Appendix A: Data QA (missingness and coverage)

```{r, results='asis'}
safe_input_table("tab_missing_overall.tex",
caption = "Missingness by variable at the country–year level.")
safe_input_table("tab_coverage_country.tex",
                 caption = "Coverage by origin country for key variables (share of years observed)."
)
```


## Appendix B: Baseline regression panels


```{r, results='asis'}
safe_input_table("tab_main_total_A.tex",
caption = "Baseline TWFE (Total): occupation shares (Blue–White).")
safe_input_table("tab_main_total_B.tex",
caption = "Baseline TWFE (Total): no-occupation share and class-of-admission shares.")
safe_input_table("tab_main_net_A.tex",
caption = "Baseline TWFE (Net): occupation shares (Blue–White), controlling for adjustment share.")
safe_input_table("tab_main_net_B.tex",
caption = "Baseline TWFE (Net): no-occupation share and class-of-admission shares, controlling for adjustment share.")
```

## Appendix C: Additional robustness tables

```{r, results='asis'}
safe_input_table("tab_spec_compare_blue.tex",
caption = "Specification comparison (blue share, net): baseline vs trends, weights, and balanced panel.")
safe_input_table("tab_blue_trends_robustness.tex",
caption = "Robustness to origin-specific linear trends (blue share).")
safe_input_table("tab_robust_extra_blue.tex",
caption = "Additional sensitivity checks (blue share, net): drop high-missing years, drop incomplete-coverage countries, and restrict window.")
safe_input_table("tab_weights_blue.tex",
caption = "Weighted specifications (blue share, net): admissions weights and occupation-count weights.")
safe_input_table("tab_balanced_blue.tex",
caption = "Balanced-panel restriction (blue share, net).")
safe_input_table(
  "tab_alt_intensity_blue.tex",
  caption = "Alternative exposure construction and Mexico exclusion (blue share, net)."
)
```


## Appendix D: Decomposition tables + log plot

```{r, results='asis'}
safe_input_table("tab_decomp_levels_A.tex",
caption = "Decomposition (levels): IRCA × Post associations on blue counts and denominators.")
safe_input_table("tab_decomp_levels_B.tex",
caption = "Decomposition (log(1+x)): proportional associations on blue counts and denominators.")
```

```{r fig-decomp-logs, fig.cap="Decomposition in log(1+x): proportional associations on numerator and denominators.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_decomp_logs_only.png")
```


## Appendix E: Timing diagnostics (interpretation check)

```{r fig-event-study, fig.cap="Event-study: exposure-by-year associations for blue-collar share (relative to 1988).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_es_blue_total.png")
```

```{r, results='asis'}
safe_input_table("tab_wald_pretrends_es_blue_total.tex",
caption = "Event-study pretrend diagnostic: Wald test of joint significance for pre-1988 coefficients.")

```

Note: p-value is extremely small; it prints as 0 due to formatting/rounding.

```{r fig-pretrend-groups, fig.cap="Pre-IRCA group means (1982–1988): blue-collar share by IRCA exposure group.", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_pretrend_groups.png")
```

```{r fig-early-late, fig.cap="Early vs late post-period associations on blue-collar share (exposure-gradient).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_early_late_blue.png")
```


Additional timing checks (linear pretrend regression and placebo fake-post tests) are available in logs/tables but omitted from the PDF for brevity.

## Appendix F: Exploratory analyses


```{r fig-mechanism-scan, fig.cap="Exposure × Post associations across outcomes (exploratory).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_mechanisms_coefplot.png")
```


## Appendix G: QA artifact (residual sanity check)

```{r fig-residuals-year, fig.cap="Mean residual by year (baseline net specification).", out.width="0.90\\linewidth", fig.pos="H"}
include_fig_safe("fig_residuals_by_year_blue.png")
```

