---
title: "When a KPI Lies"
subtitle: "A Data Analyst Case Study on Denominator Sensitivity in U.S. Immigration Admissions Data"
author: "Kai A. Sanchez P."
date: "February 2026"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{tabularray}
  - \UseTblrLibrary{booktabs}
  - \usepackage{siunitx}
  - \sisetup{detect-all}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{graphicx}
  - \usepackage{adjustbox}
  - \usepackage{capt-of}
  - \usepackage[none]{hyphenat}
  - \usepackage{pdflscape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center",
  out.width = "0.88\\linewidth",
  fig.pos = "H"
)

fig_dir <- "../outputs/figures"
tab_dir <- "../outputs/tables"

fig <- function(x) file.path(fig_dir, x)
tab <- function(x) file.path(tab_dir, x)

tab_tex <- function(name, caption = NULL){
  x <- readLines(tab(name), warn = FALSE)

  drop <- grepl("^\\\\begin\\{table\\*?\\}(\\[.*\\])?\\s*$", x) |
          grepl("^\\\\end\\{table\\*?\\}\\s*$", x) |
          grepl("^\\\\centering\\s*$", x) |
          grepl("^\\\\caption\\{", x) |
          grepl("^\\\\label\\{", x)

  x <- x[!drop]
  body <- paste(x, collapse = "\n")

  cat("\n\\par\\noindent\n")
  if (!is.null(caption)) cat("\\captionof{table}{", caption, "}\n", sep = "")

  cat("\\begingroup\\footnotesize\n")
  cat("\\begin{adjustbox}{max width=\\linewidth}\n")
  cat(body, "\n")
  cat("\\end{adjustbox}\n")
  cat("\\endgroup\n")
  cat("\\par\n\n")
}

safe_input_table <- function(name, caption = NULL){
  f <- tab(name)
  if (!file.exists(f)) {
    cat("\n\\textbf{[MISSING TABLE: ", name, "]}\\par\n", sep = "")
    return(invisible(NULL))
  }
  tab_tex(name, caption = caption)
}

include_fig_safe <- function(name){
  f <- fig(name)
  if (!file.exists(f)) {
    cat(paste0("**[MISSING FIGURE: ", name, "]**"))
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}

stopifnot(dir.exists(fig_dir), dir.exists(tab_dir))
```

# Executive Summary

This project audits a percentage-based KPI built from administrative immigration data: the recorded blue-collar share of immigrant admissions to the United States.

Using a reproducible country-year panel for 19 Latin American origin countries from 1982 to 2000, the analysis asks whether higher IRCA (Immigration Reform and Control Act) legalization exposure is associated with a meaningful decline in the measured blue-collar share during the 1989–1992 surge years.

The main result is simple and important:

1. The estimated blue-collar decline is large when the denominator uses known occupations only.

2. The same result becomes much smaller when the denominator is expanded to include records with no occupation.

3. The best interpretation is therefore denominator instability and reporting sensitivity, not clear evidence of a true collapse in blue-collar admissions.

This makes the project a data analyst case study in KPI definition, denominator auditing, missingness, and responsible interpretation of administrative data.

# Why This Works as a Data Analyst Project

This is not just a historical policy project, but also a measurement project. A reporting team could easily track a metric like “blue-collar share,” see it fall during a major policy surge, and conclude that immigrant composition changed sharply. This project shows why that can be misleading when:

- denominator definitions are unstable,

- missingness changes over time,

- administrative coverage shifts during operational surges.

The broader lesson is:

> When the denominator is unstable, a percentage KPI can tell the wrong story.

That is why this project is best framed as a case study in:

- KPI design,

- denominator sensitivity,

- administrative data quality,

- and robust interpretation under imperfect measurement.

# Business Problem

Suppose an analyst is asked:

**Did immigrant occupational composition change sharply during the IRCA surge?**

A naive answer would focus only on the net blue-collar share. A better analyst would asks:

- What exactly is in the denominator?

- Did coverage change at the same time?

- Is the share moving because the numerator changed, or because the denominator changed faster?

- Are “missing” or “no occupation” records affecting the interpretation?

That is the problem this project solves.

# Data and Exposure Design

The project builds a harmonized country-year panel from public INS (Immigration and Naturalization Service) Statistical Yearbooks. It merges four inputs:

- occupation

- class of admission

- gender/age

- adjustments versus new arrivals

The final sample covers 19 Latin American origin countries from 1982 to 2000.

## Exposure measure

Country-level IRCA exposure is defined as:

\[
\text{Exposure}=\frac{\text{IRCA legalizations (1989–1992)}}{\text{Pre-IRCA admissions (1982–1985)}}.
\]

Countries with larger legalization surges relative to their earlier admissions base are treated as more highly exposed.

## Modeling approach

The core specifications are descriptive two-way fixed effects panel regressions with:

- country fixed effects,

- year fixed effects,

- clustered standard errors by origin country.

These estimates are interpreted as descriptive exposure-gradient associations, not causal treatment effects.

# The Main Insight

The most important test in the project is a same-sample comparison between two versions of the same KPI:

- **Net denominator:** blue-collar count divided by known occupations only

- **Full denominator:** blue-collar count divided by known occupations plus no-occupation records

```{r fig-main-compare, fig.cap="Main comparison: the estimated blue-share decline is much larger under the net denominator than under the full denominator.", out.width="0.82\\linewidth"}
include_fig_safe("fig_main_net_vs_full_coef.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_MAIN_net_vs_full_denom.tex",
  caption = "Main result: blue-collar share using the net versus full denominator on the same sample."
)
```

This is the core finding. The headline decline is substantial under the net denominator, but it shrinks sharply under the fuller denominator. Because both specifications use the same sample, the contrast is not driven by changing rows. It is driven by the definition of the KPI itself.

That makes the analytical takeaway clear: the apparent drop in blue-collar share is highly sensitive to denominator construction.

# Magnitudes

The project also reports standardized effect sizes under both denominator definitions.

```{r, results='asis'}
safe_input_table(
  "tab_magnitudes_blue_NET.tex",
  caption = "Standardized effect sizes for blue-collar share using the net denominator."
)
safe_input_table(
  "tab_magnitudes_blue_FULLDEN.tex",
  caption = "Standardized effect sizes for blue-collar share using the full denominator."
)
```

These magnitudes reinforce the same message as the main comparison: the effect is much larger when the KPI uses the narrower denominator.

# Dashboard View of the Story

A good dashboard should not show only one headline KPI. It should show the surrounding context that helps interpret it correctly.

```{r fig-dashboard, fig.cap="Dashboard view of the denominator-sensitivity story. The shaded region marks the 1989–1992 IRCA window.", out.width="0.96\\linewidth"}
include_fig_safe("fig_PORTFOLIO_dashboard.png")
```

Taken together, the dashboard shows why the single headline share can be misleading. The net share, full share, no-occupation share, and denominator count do not tell the same story, and that divergence is precisely the point of the project.

# Looking Under the Hood: Numerator vs Denominator

A falling share can happen for at least two very different reasons:

- the numerator falls,

- or the denominator rises faster than the numerator.

To separate those possibilities, the project decomposes the share into underlying count components.

```{r fig-decomp-levels, fig.cap="Decomposition in levels: denominator counts rise more strongly than the blue-collar numerator.", out.width="0.90\\linewidth"}
include_fig_safe("fig_decomp_levels_only.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_decomp_levels_A.tex",
  caption = "Decomposition in levels."
)
```

The decomposition is one of the strongest parts of the project. The blue-collar numerator does not simply collapse. Instead, denominator counts expand strongly in higher-exposure settings, which makes the share fall mechanically.

That is the move that turns this from a regression project into a real analytics audit.

## Proportional version

```{r fig-decomp-logs, fig.cap="Decomposition in log(1+x): the proportional pattern also points to denominator growth.", out.width="0.90\\linewidth"}
include_fig_safe("fig_decomp_logs_only.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_decomp_levels_B.tex",
  caption = "Decomposition in log(1+x)."
)
```

The proportional specifications tell a similar story, which increases confidence that the main interpretation is not an artifact of scale alone.

# Denominator Channel

A natural denominator diagnostic is the behavior of the no-occupation category.

```{r fig-noocc-trends, fig.cap="No-occupation share over time by IRCA exposure group. This is descriptive evidence and should be interpreted jointly with the regression results.", out.width="0.90\\linewidth"}
include_fig_safe("fig_noocc_trends_by_exposure.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_noocc_did.tex",
  caption = "Regression evidence for the no-occupation share on the net sample."
)
```

This is an important nuance in the final version of the project: the descriptive trend figure and the regression evidence should be read together, not as a simple monotonic story. The safe interpretation is that denominator composition is unstable around the surge years, which is exactly why the blue-share KPI must not be interpreted in isolation.

# Robustness

The project does not need endless robustness tables, but it does need enough evidence to show that the main pattern is not a fragile one-off.

## Core robustness comparison

```{r, results='asis'}
safe_input_table(
  "tab_spec_compare_blue.tex",
  caption = "Comparison of key robustness specifications for the blue-collar share."
)
```

The result remains visible across key specifications such as country trends, weighting, and balanced-panel restrictions.

## Additional sensitivity checks

```{r, results='asis'}
safe_input_table(
  "tab_robust_extra_blue.tex",
  caption = "Additional sensitivity checks for the blue-collar share."
)
```

These checks matter because they show the signal is not entirely dependent on one exact coding choice or time window.

## Leave-one-country-out

```{r fig-loo, fig.cap="Leave-one-country-out influence on the baseline blue-share estimate.", out.width="0.88\\linewidth"}
include_fig_safe("fig_loo_blue_net.png")
```

The leave-one-country-out exercise shows that the baseline result is not driven by one single country.

## Small-cluster inference

```{r, results='asis'}
safe_input_table(
  "tab_wildboot_blue_net.tex",
  caption = "Wild cluster bootstrap inference for the baseline net-denominator specification."
)
```

Because the panel is small at the country-cluster level, the wild bootstrap is a useful inference check.

# Timing and Interpretation Limits

This project is descriptive, not causal. That is an important strength, not a weakness, because the report makes the limitation explicit rather than hiding it.

## Event-study timing check

```{r fig-event-study, fig.cap="Event-study timing check for the blue-share result. The shaded area corresponds to 1989–1992.", out.width="0.90\\linewidth"}
include_fig_safe("fig_es_blue_total.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_wald_pretrends_es_blue_total.tex",
  caption = "Wald test of joint pre-trend coefficients in the event-study specification."
)
```

The timing diagnostics show that higher-exposure countries were already evolving somewhat differently before the surge period. That is why the project avoids causal language and instead emphasizes measurement and KPI interpretation.

## Pre-period group means

```{r fig-pretrend-groups, fig.cap="Pre-IRCA group means for the blue-collar share.", out.width="0.88\\linewidth"}
include_fig_safe("fig_pretrend_groups.png")
```

# Supporting Diagnostics

## Exposure distribution

```{r fig-exposure-hist, fig.cap="Distribution of IRCA exposure across origin countries.", out.width="0.75\\linewidth"}
include_fig_safe("fig_exposure_hist.png")
```

## Exposure versus change in blue share

```{r fig-exposure-scatter, fig.cap="Country-level exposure versus change in the blue-collar share between pre-period and IRCA years.", out.width="0.90\\linewidth"}
include_fig_safe("fig_scatter_exposure_dblue_all.png")
```

These visuals help explain the structure of the sample and the variation behind the panel regressions.

# Reporting Recommendation

The most practical output of the project is a KPI recommendation. For any stakeholder-facing dashboard or monitoring workflow, report these together:

1. Blue share (net denominator)

2. Blue share (full denominator)

3. No-occupation share

4. Denominator count / coverage context

That package makes it much harder to mistake a denominator artifact for a real compositional shift.

## What I would do in production

If this KPI were deployed in a real reporting workflow, I would add simple rules such as:

- flag cases where the net share moves but the full share does not,

- flag large changes in the no-occupation share,

- require denominator counts to be reviewed before interpreting share movements,

- document the denominator definition directly in the dashboard.

# Portfolio Takeaway

This project demonstrates the ability to:

- build a reproducible panel from messy administrative-style data,

- define and operationalize a country-level exposure measure,

- audit denominator sensitivity in a headline KPI,

- decompose a share into numerator and denominator channels,

- run robustness checks without overstating causal claims,

- translate technical output into a stakeholder-facing measurement recommendation.

That is what makes it a strong data analyst portfolio project.

# How to Run

1. Put the raw Excel inputs in `data_raw/`

2. Open the project in R

3. Run `irca.R`

4. Render this `.Rmd`

Outputs are written to:

- `outputs/tables/`

- `outputs/figures/`

- `outputs/logs/`

- `data_clean/panel.rds`

# Sources

- U.S. Immigration and Naturalization Service (INS). *Statistical Yearbook of the Immigration and Naturalization Service*, various years.

# Appendix

## Appendix A: Data quality

```{r fig-missing, fig.cap="Missingness heatmap by country-year and variable.", out.width="0.96\\linewidth"}
include_fig_safe("fig_missing_heatmap.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_missing_overall.tex",
  caption = "Missingness by variable at the country-year level."
)
safe_input_table(
  "tab_coverage_country.tex",
  caption = "Coverage by origin country for key variables."
)
```

## Appendix B: Baseline regression panels

```{r, results='asis'}
safe_input_table(
  "tab_main_total_A.tex",
  caption = "Baseline TWFE (Total): occupation shares."
)
safe_input_table(
  "tab_main_total_B.tex",
  caption = "Baseline TWFE (Total): no-occupation share and class-of-admission shares."
)
safe_input_table(
  "tab_main_net_A.tex",
  caption = "Baseline TWFE (Net): occupation shares controlling for adjustment share."
)
safe_input_table(
  "tab_main_net_B.tex",
  caption = "Baseline TWFE (Net): no-occupation share and class-of-admission shares controlling for adjustment share."
)
```

## Appendix C: Additional supporting outputs

```{r fig-early-late, fig.cap="Early versus later post-period estimates for the blue-share result.", out.width="0.88\\linewidth"}
include_fig_safe("fig_early_late_blue.png")
```

```{r fig-bins, fig.cap="Estimated blue-share associations by exposure quartile.", out.width="0.75\\linewidth"}
include_fig_safe("fig_bins_blue.png")
```

```{r fig-mech, fig.cap="Mechanism scan across related outcomes.", out.width="0.88\\linewidth"}
include_fig_safe("fig_mechanisms_coefplot.png")
```

```{r, results='asis'}
safe_input_table(
  "tab_alt_intensity_blue.tex",
  caption = "Alternative exposure construction and Mexico exclusion."
)
safe_input_table(
  "tab_weights_blue.tex",
  caption = "Weighted specifications."
)
safe_input_table(
  "tab_balanced_blue.tex",
  caption = "Balanced-panel restriction."
)
safe_input_table(
  "tab_blue_trends_robustness.tex",
  caption = "Robustness to origin-specific linear trends."
)
```